---
title: "Parameter Checks For LOF, Isolation Forest, and GAD"
format:
  html:
    toc: true
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

```{r}
report_dir <- dirname(knitr::current_input(dir = TRUE))
project_dir <- normalizePath(file.path(report_dir, "..", ".."), winslash = "/", mustWork = FALSE)

if (!"package:AnomalyDetectionMethods" %in% search()) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all(project_dir)
  } else {
    stop("Install `devtools` or load `AnomalyDetectionMethods` before rendering this report.")
  }
}

demo_path <- file.path(project_dir, "inst", "extdata", "anomaly_report_demo.csv")
if (!file.exists(demo_path)) {
  demo_path <- system.file("extdata", "anomaly_report_demo.csv", package = "AnomalyDetectionMethods")
}
demo_df <- read.csv(demo_path, stringsAsFactors = FALSE)
feature_columns <- c("featureA", "featureB", "featureC")
demo_df
```

## Goal

These checks are designed to answer a practical tuning question:

- Are the flagged anomalies stable when settings change a little?
- Do different settings mostly preserve the same top-ranked rows?
- Does the Gaussian model look plausible under a given preprocessing choice?

The point is not to find a single mathematically optimal setting from unlabeled data, but to find settings that are stable, interpretable, and operationally manageable.

## LOF Parameter Grid

```{r}
lof_grid <- evaluateLofGrid(
  demo_df,
  featureColumns = feature_columns,
  kValues = c(3, 5, 7, 9),
  scoreCutoffs = c(1.2, 1.4, 1.6),
  topN = 5
)

lof_grid$summary
```

```{r}
lof_grid$stability
```

Interpretation:

- Look for `k` values that keep `topNValues` similar across nearby settings.
- Large swings in `flagCount` or low `topNJaccard` suggest the neighborhood size is too fragile.
- If a row remains in the top set across many `k` values, it is a stronger candidate anomaly.

## Isolation Forest Parameter Grid

```{r}
if_grid <- evaluateIsolationForestGrid(
  demo_df,
  featureColumns = feature_columns,
  ntreesValues = c(100, 300, 600),
  sampleSizeValues = c(8, 12, 15),
  scoreCutoffs = c(0.58, 0.62, 0.66),
  seeds = c(11, 23, 47, 89),
  topN = 5,
  weightsAsSampleProb = FALSE,
  nthreads = 1
)

if_grid$summary
```

```{r}
if_grid$stability
```

Interpretation:

- Higher `meanPairwiseSpearman` means the score ranking is more reproducible across seeds.
- Higher `meanPairwiseTopJaccard` means the top anomalies are repeating more consistently.
- Increase `ntrees` until those stability metrics stop improving materially.
- Compare `sampleSize` values to see whether the model behaves more locally or globally.

## Gaussian Anomaly Detection Grid

```{r}
gad_grid <- evaluateGadGrid(
  demo_df,
  featureColumns = feature_columns,
  transformMethods = c(NA, "weighted_zscore", "log1p"),
  covarianceMethods = c("classical", "mcd"),
  tailProbValues = c(0.95, 0.99),
  topN = 5,
  mcdAlpha = 0.9
)

gad_grid$summary
```

```{r}
gad_grid$stability
```

Interpretation:

- Higher `qqCorrelation` means the ranked squared distances align better with the theoretical chi-square pattern, which is a useful Gaussian fit diagnostic.
- If the top rows change drastically across `transformMethod` or `covarianceMethod`, then GAD is sensitive to distributional assumptions.
- If classical and robust covariance agree on the top anomalies, confidence in those flags increases.

## A Practical Selection Workflow

1. Start with a broad but plausible grid.
2. Eliminate settings with unstable top sets or highly volatile flag counts.
3. Prefer settings where nearby choices produce similar top anomalies.
4. For GAD, prefer transformations and covariance options with better QQ behavior and reasonable sensitivity.
5. Choose a cutoff based on review capacity after selecting a stable scoring configuration.

## Final Check On The Demo Data

```{r}
stable_lof <- localOutlierFactorOutliers(
  demo_df,
  featureColumns = feature_columns,
  k = 5,
  scoreCutoff = 1.4
)
stable_if <- isolationForestOutliers(
  demo_df,
  featureColumns = feature_columns,
  ntrees = 600,
  sampleSize = 12,
  scoreCutoff = 0.62,
  weightsAsSampleProb = FALSE,
  seed = 123,
  nthreads = 1
)
stable_gad <- gaussianAnomalyOutliers(
  demo_df,
  featureColumns = feature_columns,
  transformMethod = "log1p",
  covarianceMethod = "mcd",
  mcdAlpha = 0.9,
  tailProb = 0.99
)

compareAnomalyResults(
  list(
    lof = stable_lof,
    isolation_forest = stable_if,
    gaussian = stable_gad
  )
)
```
