---
title: "LOF, Isolation Forest, and Gaussian Anomaly Detection"
format:
  html:
    toc: true
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

```{r}
report_dir <- dirname(knitr::current_input(dir = TRUE))
project_dir <- normalizePath(file.path(report_dir, "..", ".."), winslash = "/", mustWork = FALSE)

if (!"package:AnomalyDetectionMethods" %in% search()) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all(project_dir)
  } else {
    stop("Install `devtools` or load `AnomalyDetectionMethods` before rendering this report.")
  }
}

demo_path <- file.path(project_dir, "inst", "extdata", "anomaly_report_demo.csv")
if (!file.exists(demo_path)) {
  demo_path <- system.file("extdata", "anomaly_report_demo.csv", package = "AnomalyDetectionMethods")
}
demo_df <- read.csv(demo_path, stringsAsFactors = FALSE)
feature_columns <- c("featureA", "featureB", "featureC")
demo_df
```

## Purpose

This dataset mixes:

- two dense clusters that represent valid but multimodal structure
- one sparse local anomaly near the first cluster
- one strong global anomaly far from both clusters
- one heavy-tailed feature pattern through `featureC`

That mix is useful for showing the practical tradeoffs:

- LOF is strong at local density departures
- Isolation Forest is strong at broad isolation-based ranking
- Gaussian anomaly detection is easy to interpret geometrically, but is more sensitive to multimodality and shape assumptions

## Fit The Models

```{r}
lof_out <- localOutlierFactorOutliers(
  demo_df,
  featureColumns = feature_columns,
  k = 5,
  scoreCutoff = 1.4
)

if_out <- isolationForestOutliers(
  demo_df,
  featureColumns = feature_columns,
  ntrees = 500,
  sampleSize = nrow(demo_df),
  scoreCutoff = 0.62,
  weightsAsSampleProb = FALSE,
  seed = 123,
  nthreads = 1
)

gad_out <- gaussianAnomalyOutliers(
  demo_df,
  featureColumns = feature_columns,
  transformMethod = "log1p",
  covarianceMethod = "classical",
  tailProb = 0.99
)

comparison <- compareAnomalyResults(
  results = list(
    lof = lof_out,
    isolation_forest = if_out,
    gaussian = gad_out
  )
)

comparison[order(-comparison$consensusCount, -comparison$lof_score), ]
```

## LOF

LOF is most interpretable as a relative score. Values near 1 look locally typical; larger values indicate local isolation compared with neighbors.

```{r}
plotAnomalyScoreDistribution(
  lof_out,
  scoreColumn = "localOutlierFactor",
  cutoffColumn = "scoreCutoff"
)
```

```{r}
plotAnomalyFeatureScatter(
  lof_out,
  xColumn = "featureA",
  yColumn = "featureB",
  scoreColumn = "localOutlierFactor"
)
```

```{r}
lof_sensitivity <- summarizeLofSensitivity(
  demo_df,
  featureColumns = feature_columns,
  kValues = c(3, 5, 7),
  scoreCutoff = 1.4
)

lof_sensitivity[, c("value", "frequency", "meanLofScore", "sdLofScore", "flagRate")]
```

## Isolation Forest

Isolation Forest is most useful as a ranking model. The score ordering is often more stable than any single threshold.

```{r}
plotAnomalyRank(
  if_out,
  scoreColumn = "isolationForestScore",
  cutoffColumn = "scoreCutoff"
)
```

```{r}
plotAnomalyPcaProjection(
  if_out,
  featureColumns = feature_columns,
  scoreColumn = "isolationForestScore",
  scale. = TRUE
)
```

```{r}
if_stability <- summarizeIsolationForestStability(
  demo_df,
  featureColumns = feature_columns,
  seeds = c(11, 23, 47, 89),
  ntrees = 300,
  sampleSize = nrow(demo_df),
  scoreCutoff = 0.62,
  weightsAsSampleProb = FALSE,
  nthreads = 1
)

if_stability[, c("value", "frequency", "meanIsolationForestScore", "sdIsolationForestScore", "flagRate")]
```

## Gaussian Anomaly Detection

GAD communicates well through Mahalanobis distance and Gaussian density, but you should always inspect whether the Gaussian geometry is plausible.

```{r}
plotAnomalyScoreDistribution(
  gad_out,
  scoreColumn = "mahalanobisDistance",
  cutoffColumn = "distanceCutoff"
)
```

```{r}
plotGadDistanceQQ(
  gad_out,
  featureColumns = feature_columns
)
```

```{r}
plotAnomalyPcaProjection(
  gad_out,
  featureColumns = feature_columns,
  scoreColumn = "mahalanobisDistance",
  scale. = TRUE
)
```

```{r}
gad_robust <- gaussianAnomalyOutliers(
  demo_df,
  featureColumns = feature_columns,
  transformMethod = "log1p",
  covarianceMethod = "mcd",
  mcdAlpha = 0.9,
  tailProb = 0.99
)

data.frame(
  value = demo_df$value,
  classical_distance = gad_out$mahalanobisDistance,
  robust_distance = gad_robust$mahalanobisDistance,
  classical_flag = gad_out$isOutlier,
  robust_flag = gad_robust$isOutlier
)
```

## Reading The Results

- Rows with high LOF but moderate Gaussian distance are often local density anomalies rather than global geometric outliers.
- Rows with high Isolation Forest score across several seeds are stable tree-based anomalies.
- Rows that remain extreme under both classical and robust GAD are strong candidates for structural outliers rather than covariance artifacts.
- Disagreement across methods is informative: it usually tells you whether the anomaly is local, global, or driven by shape assumptions.
