---
title: "Learning Plausible Measurement Ranges By Consensus Voting"
format:
  html:
    toc: true
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

```{r}
report_dir <- dirname(knitr::current_input(dir = TRUE))
project_dir <- normalizePath(file.path(report_dir, "..", ".."), winslash = "/", mustWork = FALSE)

if (!"package:AnomalyDetectionMethods" %in% search()) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all(project_dir)
  } else {
    stop("Install `devtools` or load `AnomalyDetectionMethods` before rendering this report.")
  }
}

demo_path <- file.path(project_dir, "inst", "extdata", "plausible_range_demo.csv")
if (!file.exists(demo_path)) {
  demo_path <- system.file("extdata", "plausible_range_demo.csv", package = "AnomalyDetectionMethods")
}
demo_df <- read.csv(demo_path, stringsAsFactors = FALSE)
demo_df
```

## Purpose

This example shows how to learn a plausible value range from a single
value-frequency distribution by:

1. running several univariate outlier detectors
2. converting their outputs into binary flags
3. combining those flags by consensus voting
4. deriving plausible bounds from the retained values

The demo distribution has:

- a dense central region
- thin but plausible tails
- one obvious high-end implausible value

## Run The Univariate Methods

```{r}
tukey_out <- tukeyFences(demo_df)
quantile_out <- quantileThresholds(demo_df, lowerProb = 0.01, upperProb = 0.99)
z_out <- zScoreOutliers(demo_df, zCutoff = 3)
modified_out <- modifiedZScoreOutliers(demo_df, zCutoff = 3.5)
esd_out <- generalizedESDOutliers(demo_df, maxOutliers = 5, alpha = 0.05)

comparison <- compareUnivariateOutlierResults(
  list(
    tukey = tukey_out,
    quantile = quantile_out,
    zscore = z_out,
    modified = modified_out,
    esd = esd_out
  )
)

comparison
```

## Inspect The Voting Structure

```{r}
comparison[, c(
  "value",
  "frequency",
  "tukey_flag",
  "quantile_flag",
  "zscore_flag",
  "modified_flag",
  "esd_flag",
  "consensusCount"
)]
```

Here, `consensusCount` is the number of methods that flagged each value as an outlier.

## Learn Plausible Ranges Under Different Assumptions

```{r}
consensus <- learnPlausibleRangeConsensus(
  comparisonDf = comparison,
  consensusMethod = c("count", "frequency_weighted", "support_weighted"),
  consensusThreshold = 3,
  rangeMethod = c("weighted_quantile", "raw_min_max"),
  lowerProb = 0.001,
  upperProb = 0.999
)

consensus$rangeSummary
```

## Compare Retained Values

```{r}
consensus$valueSummary
```

Useful columns:

- `consensusCount`: number of methods flagging the value
- `consensusScore_count`: plain vote count
- `consensusScore_frequency_weighted`: vote count multiplied by frequency
- `consensusScore_support_weighted`: vote count multiplied by relative support
- `isConsensusOutlier_*`: whether the value is excluded under each consensus strategy

## Reading The Differences

- `count` consensus is the closest to a simple majority-voting approach.
- `frequency_weighted` penalizes repeated flagged values more heavily.
- `support_weighted` downweights rare values, which can help preserve unusual but plausible extremes.
- `weighted_quantile` bounds are more robust than `raw_min_max` because they are less sensitive to single retained edge values.

## Practical Interpretation

For measurement-range learning, the usual workflow is:

1. start with `consensusMethod = "count"`
2. use a moderate threshold such as `3`
3. compare `weighted_quantile` against `raw_min_max`
4. inspect whether rare but clinically plausible values are being retained
5. increase or decrease the consensus threshold depending on how conservative you want the learned range to be

## Optional Next Step

If the distribution is strongly skewed, repeat the process after a log
transformation and compare the resulting plausible bounds.
