---
title: "SBP Analysis Blueprint For LOF, Isolation Forest, and GAD"
format:
  html:
    toc: true
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

```{r}
report_dir <- dirname(knitr::current_input(dir = TRUE))
project_dir <- normalizePath(file.path(report_dir, "..", ".."), winslash = "/", mustWork = FALSE)

if (!"package:AnomalyDetectionMethods" %in% search()) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all(project_dir)
  } else {
    stop("Install `devtools` or load `AnomalyDetectionMethods` before rendering this report.")
  }
}

demo_path <- file.path(project_dir, "inst", "extdata", "sbp_strata_demo.csv")
if (!file.exists(demo_path)) {
  demo_path <- system.file("extdata", "sbp_strata_demo.csv", package = "AnomalyDetectionMethods")
}
sbp_df <- read.csv(demo_path, stringsAsFactors = FALSE)
sbp_df
```

## Proposed Dataset Schema

Recommended row definition:

- one row per `SBP value x age band x gender x site`
- `value` is the SBP value you want to evaluate
- `frequency` is the number of measurements with that exact SBP in that context
- `ageBand` is the interpretable 5-year bucket
- `ageMid` is the numeric midpoint used as a feature when you want age to remain in the model
- `gender` and `site` are used for stratification rather than numeric distance calculations

In this framing, the anomaly detectors should generally use:

- `featureColumns = c("value", "ageMid")`

within a chosen `(site, gender)` stratum.

## Preprocessing Steps

1. Restrict to the intended blood-pressure measurement concept and unit.
2. Remove implausible or clearly miscoded values if you want to focus on distributional anomalies rather than impossible values.
3. Bin age into 5-year bands for interpretability and create `ageMid` for modeling.
4. Standardize site and gender values.
5. Aggregate to the row definition above.
6. Run stratum diagnostics before fitting anomaly detectors.

## Stratum Diagnostics

```{r}
strata_diag <- summarizeStrataDiagnostics(
  sbp_df,
  strataColumns = c("site", "gender"),
  valueColumn = "value",
  frequencyColumn = "frequency",
  kValues = c(3, 5, 10),
  featureColumns = "ageMid"
)

strata_diag
```

```{r}
flagged_strata <- flagAnalyzableStrata(
  strata_diag,
  minDistinctValues = 6,
  minTotalFrequency = 20,
  lofK = 5
)

flagged_strata
```

```{r}
plotStratumDiagnostics(
  flagged_strata,
  xColumn = "totalFrequency",
  yColumn = "distinctValues",
  colorColumn = "usableForLof"
)
```

How to read this:

- `distinctValues` tells you whether the stratum has enough SBP support points.
- `totalFrequency` tells you whether the stratum has enough mass to be worth modeling.
- `top1Mass` and `top5Mass` tell you whether the distribution is overly concentrated.
- `supportsK_*` tells you whether candidate LOF neighborhood sizes are even plausible.
- `usableForLof`, `usableForIf`, and `usableForGad` are first-pass readiness flags, not guarantees.

## First-Pass Modeling Workflow

For a first proof of concept, select a stratum that looks analyzable. Here we use Site A / Female.

```{r}
analysis_df <- subset(sbp_df, site == "A" & gender == "F")
analysis_df
```

Use SBP value itself plus age midpoint as the numeric feature space:

```{r}
feature_columns <- c("value", "ageMid")
```

### LOF

```{r}
lof_out <- localOutlierFactorOutliers(
  analysis_df,
  featureColumns = feature_columns,
  k = 3,
  scoreCutoff = 1.3
)

lof_out
```

### Isolation Forest

```{r}
if_out <- isolationForestOutliers(
  analysis_df,
  featureColumns = feature_columns,
  ntrees = 300,
  sampleSize = nrow(analysis_df),
  scoreCutoff = 0.6,
  weightsAsSampleProb = FALSE,
  seed = 123,
  nthreads = 1
)

if_out
```

### Gaussian Anomaly Detection

```{r}
gad_out <- gaussianAnomalyOutliers(
  analysis_df,
  featureColumns = feature_columns,
  transformMethod = "weighted_zscore",
  covarianceMethod = "classical",
  tailProb = 0.99
)

gad_out
```

## Compare The First-Pass Results

```{r}
compareAnomalyResults(
  list(
    lof = lof_out,
    isolation_forest = if_out,
    gaussian = gad_out
  )
)
```

## Recommended Workflow For Real SBP Data

1. Build the aggregated SBP dataset at `value x age band x gender x site`.
2. Run `summarizeStrataDiagnostics()` across `(site, gender)` strata.
3. Use `flagAnalyzableStrata()` to identify strata that are large and varied enough to support modeling.
4. Within usable strata, run LOF, IF, and GAD on `c("value", "ageMid")`.
5. Use the existing tuning helpers to assess stability of:
   - LOF across `k`
   - IF across `ntrees`, `sampleSize`, and seeds
   - GAD across preprocessing and covariance choices
6. Prioritize rows flagged consistently across methods or consistently across nearby settings.

## Practical Caution

If you stratify by `site`, `gender`, and `ageBand` all at once, then age is no longer a varying feature within the stratum. In that case, the models are effectively working on SBP values alone. If you want age to remain a feature, stratify by `site` and `gender`, and let `ageMid` vary within that stratum.
